Analysis of cloud enhancement events in a 30-year record of global solar irradiance at Thessaloniki, Greece

### 2024-09-09 article/article.tex
Commit 2024-09-09 21:18

  \begin{abstract}

  \end{abstract}
    \begin{keyword}
    cloud enhancement \sep total solar radiation \sep global horizontal
irradiance \sep 
    over irradiance
  \end{keyword}

@@

\hypertarget{abstract}{%
\section*{Abstract}\label{abstract}}
\addcontentsline{toc}{section}{Abstract}

\textbf{Abstract temporary placed here for technical reasons}

In this study, we investigate the characteristics of global horizontal
irradiance enhancement events induced by clouds over Thessaloniki for
the period 1994 -- 2023 using data recorded every one minute. To our
knowledge, there isn't a similar study with such a long time series. We
identified the cloud enhancement (CE) events by creating an appropriate
cloud-free irradiance reference using a radiative transfer model and
aerosol optical depth data from a collocated Cimel sun photometer and a
Brewer spectrophotometer. We found a trend in CE events of
\(+112\pm 35\,\text{cases}/\text{year}\), and a trend in the
corresponding irradiation of \(+329.9\pm 112\,\text{kJ}/\text{year}\).
The peak of the CE events was observed during May and June. CE events
with duration longer than 10\nobreakspace{}min are very rare
(\(<8\,\%\)), with exceptions lasting over an hour and up to 140
minutes. Finally, we have detected enhancements above the total solar
irradiance at the top of the atmosphere for the same solar zenith angle
of up to \(204\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of the
cases below \(40\,\text{W}/\text{m}^{2}\). Most of these extreme events
occur in spring -- early summer, with a secondary peak in autumn.

@@

In addition, we provide in Figure\nobreakspace{}\ref{fig:example-year}
an example scatter plot between the measured GHI and the modeled
cloud-free irradiance for one year (2005), where the CE and ECE events
are clearly grouped above the cloud-free irradiance (approximated with
the green line). The grey colored data points above this line correspond
to the ambiguous data points discussed above. Finally, the black belt
just below the green line is formed from data measured mainly under
cloud-free conditions,conditions with the spread likely caused by the
short-term variability of AOD. A small fraction of these data
corresponds also to cases with thin cirrus clouds causing weak
attenuation of GHI, often indistinguishable from the attenuation by
aerosols.

@@

\caption{Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one-minute steps. The red line shows the modelled threshold for the detection of CE events, which are denoted with red circles. The black line represents the TOA TSI on a horizontal plane, equivalent to the threshold for the identification of ECE events, shown with red circles. The purple line is the modeled cloud-free irradiance. The dark yellow line is the solar constant of $1367\,\text{W}/\text{m}^{2}$.}\label{fig:example-day}
\end{figure}

@@

\caption{Example scatter plot of the measured GHI and the reference cloud-free irradiance in Thessaloniki for the year 2005. The over-irradiance for CE and ECE events is colour coded, while the remaining data points are shown in black. The reference green line goes through the origin with a slope of one.}\label{fig:example-year}
\end{figure}

@@

Following the application of the above discussed methodology to the
entire dataset (\(6\) million of one-minute GHI measurements),
\(2.32\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(369.3\,\text{W}/\text{m}^2\) or
\(35.3\,\%\) above the threshold. The strongest CE event of \(53\,\%\)
above the cloud-free threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI value of \(861.8\,\text{W}/\text{m}^2\) and
a OI of \(298.4\,\text{W}/\text{m}^2\). Both cases are also ECE events
with \(131.2\,\text{W}/\text{m}^2\) and \(161.5\,\text{W}/\text{m}^2\)
above the TSI at TOA for the same SZA, respectively. In the following
sections we are discussing the long-term trends and variability of the
CE events as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1994 -- 2023,
together with corresponding linear trends. To account for missing data,
all three quantities have been divided with the fraction of the valid
GHI observations in each month. Statistically significant (at the
\(95\,\%\) confidence level) increasing trends appear for the yearly
number of CE events (\(+112\pm 35\,\text{cases}/\text{year}\)) and the
excess irradiation (\(+329.9 \pm 112\,\text{kJ}/\text{year}\)), while
the the trend in the yearly mean OI is negligible
(\(+0.11\pm 0.2\,\text{W}/\text{m}^2/\text{year}\)) and of no
statistical significance. The average OI for the entire period is
\(+42.6\pm 2.8\,\text{W}/\text{m}^2\). The yearly excess irradiation due
to CE events ranges between about \(8\) and
\(24\,\text{MJ}/\text{m}^2\). On average, it is about one half of the
highest daily irradiation of \textasciitilde31 MJ/m\textsuperscript{2}
under cloud-free conditions recorded at the summer solstice in
Thessaloniki. Therefore, from solar energy perspective the yearly excess
irradiation is only a tiny fraction of the available solar energy.
Although the interannual variability of the mean OI is rather weak, the
variability about the trend lines is quite large for the number of CE
cases and the excess irradiation (panels b and c). For these two
quantities the spread tends to increase with time, suggesting a
significant variability in cloud patterns over the area, possibly
associated to changes in climate.

We must note that the excess irradiation from the CE events that is
received at the surface does not affect the balance of the energy budget
in the atmosphere. The incoming solar radiation in the region is not
increased by the OI but is rather redistributed through the CE events.
This is also depicted by the ECE irradiance values, which exceed the
equivalent cloud-free irradiance by a significant amount.

@@

Next, we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the
monthly box and whisker plot of the number of occurrences of the CE
events in percent, calculated relative to the available one-minute data
in each month. Although CE events are present throughout the year,
showing a clear seasonal cycle, the most active months are May and June,
with their median representing about \(3\,\%\) of the data. The least
frequent events (about \(1\,\%\)) occur in August and during winter
(December -- February). This seasonality is a combined effect of
different factors, among them the types of clouds, their frequency of
occurrence, the seasonally varying relative position of the sun, as well
as the local landscape characteristics that may influence the formation
of the clouds. For example, during May and June the frequent formation
of cumulus clouds in the area leads to a significant increase in the CE
events. However, the lack of detailed data in the area of cloud
characteristics does not allow further analysis. Finally, the
interannual variability of the monthly CE events is quite high as
manifested by the size of the boxes and the large monthly extremes,
especially in the summer. Investigation of the causes of this
variability is beyond the scope of this study. It could possibly be
related to the cloud cover variability in winter and autumn over
southern Europe, associated with the North Atlantic Oscillation (NAO)
circulation \citep{Chiacchio2010}, or to the observed decreasing trend
in cloud cover as a result of global warming
\citep[e.g.,][]{SanchezLorenzo2017}.

@@

The distribution of the number of CE events as a function of OI is shown
in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an inverse
relationship between the frequency of CE events and OI with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favouring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OI is
below the long-term average of \(42.6\,\text{W}/\text{m}^2\), while
\(9.6\,\%\) of the events correspond to OI larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(437\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OI
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean,
resulting in different cloud patterns, that may affect the
characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. To study the characteristics of these
consecutive events, we grouped them into bins of increasing duration in
steps of one minute. We have identified 34849 groups of CE events in the
whole period of study, where the group with the longest duration of 140
minutes occurred on 07~July 2013 with SZA ranging between \(52.1^\circ\)
and \(77.9^\circ\).
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the
frequency distribution of the CE events according to their duration. We
conclude that although some groups of events last for more than an hour,
about \(79\,\%\) have a duration of equal or less than 5 minutes.

@@

By creating a cloud-free approximation representing the long- and
short-term variability of the expected cloud-free GHI, we were able to
identify cases of CE events in Thessaloniki for the period 1994 -- 2023.
After analyzing the CE cases, we found an increase of
\(+112.1\pm 34.9\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+329.9\pm 112\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(79\,\%\) of them is equal or
bellow \(5\) minutes.

We have observed ECE cases with GHI exceeding the TOA TSI on horizontal
plane by more than \(200\,\text{W}/\text{m}^{2}\) while for \(75\,\%\)
of the cases the excess irradiance (relative to TOA TSI on horizontal
plane) is below \(40\,\text{W}/\text{m}^{2}\). The climatological
characteristics of the ECE events showed that the most active months are
spread over half of the year, and particularly in the periods March --
June and September -- October. The magnitude of the ECE events
identified in Thessaloniki events does not exceed the values reported
for sites with more favourable conditions for the phenomenon
\citep[e.g.,][]{Cordero2023}. Some of the characteristics of CE and ECE
events we analyzed have strong similarities with the results of
\citet{Vamvakas2020} for the city of Patras, south of Thessaloniki,
albeit with differences in the magnitude of OI.

An interpretation of the CE trends shows that the interaction of GHI
with the clouds, through this 30-year period, is a dynamic phenomenon
that needs further investigation. For such studies, it would be
essential to have more detailed information on cloud characteristics,
especially in order to investigate possible associations of the observed
trends with changes in climate.

>>>>

  \begin{abstract}
  In this study, we investigate the characteristics of global horizontal
  irradiance enhancement events induced by clouds over Thessaloniki for
  the period 1994 -- 2023 using data recorded every one minute. To our
  knowledge, there isn't a similar study with such a long time series.
  We identified the cloud enhancement (CE) events by creating an
  appropriate cloud-free irradiance reference using a radiative transfer
  model and aerosol optical depth data from a collocated Cimel sun
  photometer and a Brewer spectrophotometer. We found a trend in CE
  events of \(+112\pm 35\,\text{cases}/\text{year}\), and a trend in the
  corresponding irradiation of \(+329.9\pm 112\,\text{kJ}/\text{year}\).
  The peak of the CE events was observed during May and June. CE events
  with duration longer than 10\nobreakspace{}min are very rare
  (\(<8\,\%\)), with exceptions lasting over an hour and up to 140
  minutes. Finally, we have detected enhancements above the total solar
  irradiance at the top of the atmosphere for the same solar zenith
  angle of up to \(204\,\text{W}/\text{m}^{2}\), with the \(75\,\%\) of
  the cases below \(40\,\text{W}/\text{m}^{2}\). Most of these extreme
  events occur in spring -- early summer, with a secondary peak in
  autumn.
  \end{abstract}
    \begin{keyword}
    cloud enhancement \sep total solar radiation \sep global horizontal
irradiance \sep 
    over irradiance
  \end{keyword}

@@

@@

In addition, we provide in Figure\nobreakspace{}\ref{fig:example-year}
an example scatter plot between the measured GHI and the modeled
cloud-free irradiance for year 2005, where the CE and ECE events are
clearly grouped above the cloud-free irradiance (approximated with the
green line). The gray colored data points above this line correspond to
the ambiguous data points discussed above. Finally, the black belt just
below the green line is formed from data measured mainly under
cloud-free conditions, with the spread likely caused by the short-term
variability of AOD. A small fraction of these data corresponds also to
cases with thin cirrus clouds causing weak attenuation of GHI, often
indistinguishable from the attenuation by aerosols.

@@

\caption{Example of CE identification in Thessaloniki for 2019-07-11. The green line with blue symbols depicts the measured GHI in one-minute steps. The red line shows the modeled threshold for the detection of CE events, which are denoted with gray circles. The black line represents the TOA TSI on a horizontal plane, equivalent to the threshold for the identification of ECE events, shown with red circles. The purple line is the modeled cloud-free irradiance. The dark yellow line is the solar constant of $1367\,\text{W}/\text{m}^{2}$. The three sky images below the figure are taken at three characteristic instances (a, b, c) denoted on the figure with vertical dashed lines.}\label{fig:example-day}
\end{figure}

@@

\caption{Example scatter plot of the measured GHI and the reference cloud-free irradiance in Thessaloniki for the year 2005. The over-irradiance for CE and ECE events is color coded, while the remaining data points are shown in black. The reference green line goes through the origin with a slope of unity.}\label{fig:example-year}
\end{figure}

@@

Following the application of the above discussed methodology to the
entire dataset (\(>6\) million of one-minute GHI measurements),
\(2.32\,\%\) were identified as CE events and \(0.036\,\%\) as ECE
events. The highest recorded GHI due to CE was
\(1416.6\,\text{W}/\text{m}^2\) on 24~May 2007 at a SZA of
\(19.9^\circ\) corresponding to OI of \(369.3\,\text{W}/\text{m}^2\) or
\(35.3\,\%\) above the threshold. The strongest CE event of \(53\,\%\)
above the cloud-free threshold was observed on 28~October 2016 at a SZA
of \(59.2^\circ\) with a GHI of \(861.8\,\text{W}/\text{m}^2\) and a OI
of \(298.4\,\text{W}/\text{m}^2\). Both cases are also ECE events with
\(131.2\,\text{W}/\text{m}^2\) and \(161.5\,\text{W}/\text{m}^2\) above
the TSI at TOA for the same SZA, respectively. In the following sections
we are discussing the long-term trends and variability of the CE events
as well as of the corresponding OI and excess irradiation.

@@

Figure\nobreakspace{}\ref{fig:P-energy} shows the time series of the
yearly number of CE cases (each with duration of one minute), the yearly
mean OI and the yearly excess irradiation for the period 1994 -- 2023,
together with corresponding linear trends. To account for missing data,
all three quantities have been divided with the fraction of the valid
GHI observations in each month. Statistically significant (at the
\(95\,\%\) confidence level) increasing trends appear for the yearly
number of CE events (\(+112\pm 35\,\text{cases}/\text{year}\)) and the
excess irradiation (\(+329.9 \pm 112\,\text{kJ}/\text{year}\)), while
the trend in the yearly mean OI is negligible
(\(+0.1\pm 0.2\,\text{W}/\text{m}^2/\text{year}\)) and of no statistical
significance. The average OI for the entire period is
\(+42.6\pm 2.8\,\text{W}/\text{m}^2\). The yearly excess irradiation due
to CE events ranges between about \(8\) and
\(24\,\text{MJ}/\text{m}^2\). On average, it is about one half of the
highest daily irradiation of \textasciitilde31 MJ/m\textsuperscript{2}
under cloud-free conditions recorded at the summer solstice in
Thessaloniki. Therefore, from solar energy perspective the yearly excess
irradiation is only a tiny fraction of the available solar energy.
Although the interannual variability of the mean OI is rather weak, the
variability about the trend lines is quite large for the number of CE
cases and the excess irradiation (panels b and c). For these two
quantities the spread tends to increase with time, suggesting a
significant variability in cloud patterns over the area, possibly
associated with changes in climate.

We must note that the excess irradiation from the CE events that is
received at the surface does not affect the balance of the energy budget
in the atmosphere. The incoming solar radiation in the region is not
increased by the OI but is rather redistributed at the specific location
as additional scattered radiation. This is also depicted by the ECE
irradiance values, which exceed the equivalent cloud-free irradiance by
a significant amount.

@@

Next, we investigated the distribution of the CE events within the year.
Figure\nobreakspace{}\ref{fig:relative-month-occurrences} shows the
monthly box and whisker plot of the number of occurrences of the CE
events in percent, calculated relative to the available one-minute data
in each month. Although CE events are present throughout the year,
showing a clear seasonal cycle, the most active months are May and June,
with their median representing about \(3\,\%\) of the data. The least
frequent events (about \(1\,\%\)) occur in August and during winter
(December -- February). This seasonality is a combined effect of
different factors, among them the types of clouds, their frequency of
occurrence, the seasonally varying relative position of the sun, as well
as the local landscape characteristics that may influence the formation
of the clouds. For example, during May and June the frequent formation
of cumulus clouds in the area leads to a significant increase in the CE
events. However, the lack of detailed data on cloud characteristics in
the area does not allow further analysis. Finally, the interannual
variability of the monthly CE events is quite high as manifested by the
size of the boxes and the large monthly extremes, especially in the
summer. Investigation of the causes of this variability is beyond the
scope of this study. It could possibly be related to the cloud cover
variability in winter and autumn over southern Europe, associated with
the North Atlantic Oscillation (NAO) circulation \citep{Chiacchio2010},
or to the observed decreasing trend in cloud cover as a result of global
warming \citep[e.g.,][]{SanchezLorenzo2017}.

@@

The distribution of the number of CE events as a function of OI is shown
in Figure\nobreakspace{}\ref{fig:ovir-distribution}. There is an inverse
relationship between the frequency of CE events and OI with an
exponential-like decline. This is expected, as the stronger the CE
events are, the rarer the conditions favoring the occurrence of CE
events. For the majority (over \(62\,\%\)) of the CE events the OI is
below the long-term average of \(42.6\,\text{W}/\text{m}^2\), while
\(9.6\,\%\) of the events correspond to OI larger than
\(100\,\text{W}/\text{m}^2\) and up to the highest value of
\(437\,\text{W}/\text{m}^2\). This distribution is indicative of the
magnitude and the probability of the expected CE events over
Thessaloniki. Similar distribution of CE events, albeit with larger OI
values, has been reported by \citet{Vamvakas2020}, for the city of
Patras. This site is located \(2.5^\circ\) south of Thessaloniki and is
exposed to air masses coming mainly from the eastern Mediterranean,
resulting in different cloud patterns that may affect the
characteristics and magnitude of the CE events.

@@

The duration of the CE events is variable and can last for several
minutes or even more than an hour. To study the characteristics of these
consecutive events, we grouped them into bins of increasing duration in
steps of one minute. We have identified 34849 groups of CE events in the
whole period of study, where the group with the longest duration of 140
minutes occurred on 07~July 2013 with SZA ranging between \(52.1^\circ\)
and \(77.9^\circ\).
Figure\nobreakspace{}\ref{fig:ceg-duration-distribution} shows the
frequency distribution of the CE events according to their duration. We
conclude that although some groups of events last for more than an hour,
about \(79\,\%\) have a duration of equal or less than 5 minutes, likely
due to the movements of clouds causing frequent blocking of the direct
irradiance, which reduce the duration of the CE events.

@@

By creating a cloud-free approximation representing the long- and
short-term variability of the expected cloud-free GHI, we were able to
identify cases of CE and ECE events in Thessaloniki for the period 1994
-- 2023. After analyzing the CE cases, we found an increasing tendency
of \(+112.1\pm 34.9\,\text{cases}/\text{year}\), with the mean annual
irradiation of the CE events increasing with a rate of
\(+329.9\pm 112\,\text{kJ}/\text{year}\). The most active months of CE
events are May and June. We found that continuous CE events can last up
to \(140\) minutes, while the duration of \(79\,\%\) of them is equal or
bellow \(5\) minutes.

We have observed ECE cases with GHI exceeding the TOA TSI on a
horizontal plane by more than \(200\,\text{W}/\text{m}^{2}\) while for
\(75\,\%\) of the cases the excess irradiance (relative to TOA TSI on a
horizontal plane) is below \(40\,\text{W}/\text{m}^{2}\). The
climatological characteristics of the ECE events showed that the most
active months are spread over half of the year, and particularly in the
months January -- March and November, while ECE events are practically
absent in July and August. The magnitude of the ECE events identified in
Thessaloniki does not exceed the values reported for sites with more
favorable conditions for the phenomenon \citep[e.g.,][]{Cordero2023}.
Some of the characteristics of CE and ECE events we analyzed have strong
similarities with the results of \citet{Vamvakas2020} for the city of
Patras, south of Thessaloniki, albeit with differences in the magnitude
of OI.

The investigation of the CE trends in Thessaloniki through this 30-year
period shows that the interaction of GHI with clouds is a complex
phenomenon that merits further attention. For such studies, it would be
essential to have more detailed information on cloud characteristics,
especially in order to investigate possible associations of the observed
trends with changes in climate. Finally, automated procedures to analyze
the sky-camera images would greatly improve our knowledge about the
generation of these events.
